Implementation of a lattice Boltzmann method for immiscible two-phase flow simulations using the level set method

### 2015-11-14 Paper/paper.tex
Kleinigkeit

For the coupling, we use a bounce back type boundary condition that prescribes the velocity as Dirichlet value as described in \cite[p. 1146]{Thoemmes}. Consider a point $x_2 \in \Omega_2$ that is next to the phase interface, where the link $c_i$ crosses the interface, see \cref{coupling}. We set,
\begin{figure}
	\hfill\includegraphics[width=8cm,natwidth=437,natheight=178]{coupling.png}\hspace*{\fill}
	\caption{Link between two points in different fluids.}
	\label{coupling}
\end{figure}
\begin{equation}
	f_i(x_2,t+1) = f_{i*}^+(x_2,t) + 6f_i^*c_i \cdot \tilde{u} + R_i \text{.}
\end{equation}
Here, $\tilde{u}$ describes the linear interpolation of the velocity along the direction $c_i$, evaluated at the location $\tilde{x} = x_1 + qc_i = x_2 + (q-1)c_i$ on the interface,
$$
\tilde{u} = qu(x_2,t) + (1-q)u(x_1,t) \text{.}
$$
The additional term $R_i$ ensures the jump conditions of the normal stress and corrects the error that is introduced by the first order accurate bounce back treatment,
\begin{equation}
  R_i = 6f_i^* \left(q(q-1)\Lambda_i:[S] - (q-\frac12)\Lambda_i:S(x_2) \right) \text{,}
  \label{R_i}
\end{equation}
with
\begin{equation}
  \Lambda_i = c_i \otimes c_i - \frac12 (c_i \cdot c_i) \mathbf{I} \text{. }
\end{equation}
That is an orthogonal projection of the column vectors of the (symmetric) deviatoric shear rate jump tensor on the link direction. This projection is weighted with the interface distance and a compensation term for the interpolation error. This error arises, when the interface is not on the exact center between two LBM grid cells. The tensor product is $a \otimes b = ab^T$ and the double contraction $A:B = \mathrm{trace}(AB^T)$. Furthermore, we use the following first-order estimate to obtain the shear rate from the non-equilibrium parts of the distributions, see also \cite{Krueger},
\begin{equation}
  S_k = -\frac{3}{2\tau } \sum_{i=1}^9 c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{.}
\end{equation}
Now we have,
\begin{equation}
  \Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - \frac12 (c_i \cdot c_i)) + 2([S]:n \otimes t)(n \cdot c_i)(t \cdot c_i) \text{,}
  \label{Lambda_iTimes_S}
\end{equation}
which is obtained by a coordinate transformation of (\ref{R_i}) onto a basis $(n, t)$ of normal and tangential vector to the interface. From the fluid mechanical jump condition (\ref{jumpconditions}) and the general relation $[\mu S] = \bar{\mu}[S] + [\mu]\bar{S} $ it can be seen,
\begin{align}
  [S]:n \otimes n &=  \frac{1}{2 \bar\mu} ([p] + 2 \sigma \kappa) - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes n \text{,} \\
  [S]:n \otimes t &= - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes t \text{.}
\end{align}
Here $\bar{\mu} = (\mu_1 + \mu_2)/2$ and $\bar{S} = (S_1 + S_2)/2$ are averaged quantities, $\kappa$ the interface curvature, and  $[\mu] = \mu_2 - \mu_1$ the viscosity jump.

@@

\subsection{Couette channel flow}
\begin{figure}[b!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette1.pdf}\hspace*{\fill}
	\caption{Two fluid domains $\Omega_i$ separated by an interface.}
	\label{couette1}
\end{figure}
The Couette channel flow consists of a lower wall $y=0$ at rest and an upper wall $y=1$ that is moving tangentially with a prescribed constant velocity. At these walls no-slip boundary conditions hold. Along the $x$-direction periodic boundary conditions are prescribed. The two fluids layers are oriented such that their interface is parallel to the no-slip walls, and no external pressure gradient is applied.
For this flow scenario an analytic solution can be derived from the Navier-Stokes equations, namely they simplify to $\frac{\partial }{\partial y}\left( \eta\frac{\partial u}{\partial y}\right) =0$. In combination with the no-slip conditions at top and bottom and wall this means, that the horizontal velocity has a linear profile with a kink at the fluid interface. The respective slopes between both phases are related by $\frac{\partial_y u^{(1)}}{\partial_y u^{(2)}} = \frac{\eta^{(2)}}{\eta^{(1)}}$.

@@

Figure \ref{couette2} shows a second scenario with $5 \times 10$ cells along $x \times y$ direction and viscosities $\mu^{(1)} = 1$ and $\mu^{(2)} = 0.2$ with the interface located at $y=0.75$. Again the numerically approximated values agree very well with the analytic ones and the simulation converged at a relative error of $1.4\cdot10^{-4}$ after 2000 LBM iterations.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette2.pdf}\hspace*{\fill}
	\caption{Two fluid domains $\Omega_i$ separated by an interface.}
	\label{couette2}
\end{figure}
We conclude from this that the implemented LBM boundary conditions successfully model viscosity jumps at the fluid interface.
\subsection{Young-Laplace experiment for a bubble}
The Couette-Flow is not influenced by surface tension due to the fact that the surface tension is multiplied with the curvature in Equation \ref{Lambda_iTimes_S} and the curvature is zero in the Couette-Flow. We use the Young-Laplace experiment to validate the implementation of surface tension. The pressure $p$ inside a bubble of radius $r$ is proportional to the surface tension coefficient $\sigma$ according to the Young-Laplace law,
\begin{equation}
	p = \frac{2\sigma}{r} = 2 \sigma \kappa .
\end{equation}

>>>>

For the coupling, we use a bounce back type boundary condition that prescribes the velocity as Dirichlet value as described in \cite[p. 1146]{Thoemmes}. Consider a point $x_2 \in \Omega_2$ that is next to the phase interface, where the link $c_i$ crosses the interface, see \cref{coupling}. We set,
\begin{figure}
	\hfill\includegraphics[width=8cm,natwidth=437,natheight=178]{coupling.png}\hspace*{\fill}
	\caption{Link between two points in different fluids.}
	\label{coupling}
\end{figure}
\begin{equation}
	f_i(x_2,t+1) = f_{i*}^+(x_2,t) + 6f_i^*c_i \cdot \tilde{u} + R_i \text{.}
\end{equation}
Here, $\tilde{u}$ describes the linear interpolation of the velocity along the direction $c_i$, evaluated at the location $\tilde{x} = x_1 + qc_i = x_2 + (q-1)c_i$ on the interface,
$$
\tilde{u} = qu(x_2,t) + (1-q)u(x_1,t) \text{.}
$$
The additional term $R_i$ ensures the jump conditions of the normal stress and corrects the error that is introduced by the first order accurate bounce back treatment,
\begin{equation}
  R_i = 6f_i^* \left(q(q-1)\Lambda_i:[S] - (q-\frac12)\Lambda_i:S_2 \right) \text{,}
  \label{R_i}
\end{equation}
with
\begin{equation}
  \Lambda_i = c_i \otimes c_i - \frac12 (c_i \cdot c_i) \mathbf{I} \text{. }
\end{equation}
That is an orthogonal projection of the column vectors of the (symmetric) deviatoric shear rate jump tensor on the link direction. This projection is weighted with the interface distance and a compensation term for the interpolation error. This error arises, when the interface is not on the exact center between two LBM grid cells. The tensor product is $a \otimes b = ab^T$ and the double contraction $A:B = \mathrm{trace}(AB^T)$. Furthermore, we use the following first-order estimate to obtain the shear rate from the non-equilibrium parts of the distributions, see also \cite{Krueger},
\begin{equation}
  S_k = -\frac{3}{2\tau } \sum_{i=1}^9 c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{.}
\end{equation}
Now we have,
\begin{equation}
  \Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - \frac12 (c_i \cdot c_i)) + 2([S]:n \otimes t)(n \cdot c_i)(t \cdot c_i) \text{,}
  \label{Lambda_iTimes_S}
\end{equation}
which is obtained by a coordinate transformation of (\ref{R_i}) onto a basis $(n, t)$ of normal and tangential vector to the interface. From the fluid mechanical jump condition (\ref{jumpconditions}) and the general relation $[\mu S] = \bar{\mu}[S] + [\mu]\bar{S} $ it can be seen,
\begin{align}
  [S]:n \otimes n &=  \frac{1}{2 \bar\mu} ([p] + 2 \sigma \kappa) - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes n \text{,} \\
  [S]:n \otimes t &= - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes t \text{.}
\end{align}
Here $\bar{\mu} = (\mu_1 + \mu_2)/2$ and $\bar{S} = (S_1 + S_2)/2$ are averaged quantities, $\kappa$ the interface curvature, and  $[\mu] = \mu_2 - \mu_1$ the viscosity jump.

@@

\subsection{Couette channel flow}
\begin{figure}[b!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette1.pdf}\hspace*{\fill}
  \caption{Two horizontal fluid layers $\Omega_i$ separated by an interface. The top boundary is moving horizontally, while the lower one is at rest}
	\label{couette1}
\end{figure}
The Couette channel flow consists of a lower wall $y=0$ at rest and an upper wall $y=1$ that is moving tangentially with a prescribed constant velocity. At these walls no-slip boundary conditions hold. Along the $x$-direction periodic boundary conditions are prescribed. The two fluids layers are oriented such that their interface is parallel to the no-slip walls, and no external pressure gradient is applied.
For this flow scenario an analytic solution can be derived from the Navier-Stokes equations, namely they simplify to $\frac{\partial }{\partial y}\left( \eta\frac{\partial u}{\partial y}\right) =0$. In combination with the no-slip conditions at top and bottom and wall this means, that the horizontal velocity has a linear profile with a kink at the fluid interface. The respective slopes between both phases are related by $\frac{\partial_y u^{(1)}}{\partial_y u^{(2)}} = \frac{\eta^{(2)}}{\eta^{(1)}}$.

@@

Figure \ref{couette2} shows a second scenario with $5 \times 10$ cells along $x \times y$ direction and viscosities $\mu^{(1)} = 1$ and $\mu^{(2)} = 0.2$ with the interface located at $y=0.75$. Again the numerically approximated values agree very well with the analytic ones and the simulation converged at a relative error of $1.4\cdot10^{-4}$ after 2000 LBM iterations.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette2.pdf}\hspace*{\fill}
	\caption{Two fluid domains $\Omega_i$ separated by an interface.}
	\label{couette2}
\end{figure}
We conclude from this that the implemented LBM boundary conditions successfully model viscosity jumps at the fluid interface.
\subsection{Young-Laplace experiment for a bubble}
Our Couette flow scenario does not account for the influence of surface tension as the fluid interface is not curved. If it would get distorded, surface tension would have a stabilizing effect although. 
Hence, we use the Young-Laplace law for the pressure drop across curved surfaces, to validate the implementation of surface tension. The pressure inside of a convex surfaces is higher than on the outside and the pressure drop $\Delta p$ of a bubble of radius $r$ is proportional to the surface tension coefficient $\sigma$ according to thefollowing relationship,
\begin{equation}
	\Delta p = \frac{2\sigma}{r} = 2 \sigma \kappa .
\end{equation}
